---
title: "Population projections using different mortality and fertility schedules"
output: html_document
---

```{r setup, include=FALSE}
# run this chunk first!
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

In this notebook, we will be exploring some of the concepts on population projection and stable population theory covered in class. 

## Section I: Introduction to the data

The aim of the first section of this lab is to introduce you to the dataset that we'll be using.

The dataset has values for $_nL_x$ (person-years lived between ages $x$ and $x+n$), $_nF_x$ (age-specific fertility) and $_nK_x$ (population between ages $x$ and $x+n$) for six countries, which have different types of mortality, fertility and age structures:

- United States
- S. Korea
- Kenya 
- Niger 

Data are for females in 2015-2020 (nLx 2019, nKx 2020, nFx 2015-2020) and were obtained from the [World Health Organisation](http://apps.who.int/gho/data/node.main.687?lang=en) ($_nL_x$) and [United Nations Population Division](https://esa.un.org/unpd/wpp/) ($_nF_x$ and $_nK_x$). 

```{r}
d <- read_csv("projection_data.csv")
```

Throughout this lab, we'll be making use of the `%>%` operator. This is like a 'pipe', which takes whatever is on the left hand side and feeds it forward into a function. For example, below we are taking our data `d`, filtering it to only include Australia, and then printing the head (the first six rows).

```{r}
d %>% 
  filter(country=="United States") %>% 
  head
```

Notice that:

- The magnitude of $_nL_x$ values suggests a radix of 100,000
- $_nF_x$ is the age-specific birth rate per 1,000 women
- $_nK_x$ is age-specific population in 000s

To save trouble later, let's change $_nL_x$ to have a radix of 1, and $_nF_x$ to have units births per woman. 

```{r}
d <- d %>% 
  mutate(nLx = nLx/10^5, 
         nFx = nFx/1000)
```


### Person-years lived: $_nL_x$

Let's investigate the values of $_nL_x$ by country. First, let's calculate the life expectancy at birth, $e_0$, for all countries. Remember that $$e_0 = \frac{\sum {_nL_x}}{l_0}$$

To calculate in R we can use `group_by` to do the calculation separately for each country.

```{r}
# calculate life expectancy at birth 
# we know radix is 1
radix <- 1

d %>% 
  group_by(country) %>% 
  summarise(e0 = sum(nLx)/radix) %>%
  arrange(e0) # arrange by increasing life expectancy
```

There is a 14 year difference in life expectancy in Niger compared to the US and a 20 year difference between Niger and S Korea. S Korea has over 7 years on the US. 

### Fertility rates: $_nF_x$

We can use the age-specific fertility rates to calculate some summary fertility measures for each country. Note that we will assume that the proportion of babies that are female (fraction of females at birth) is 0.4886. Using the information we have, we can calculate the total fertility rate (TFR), gross reproduction ratio (GRR) and net reproduction ratio (NRR). We can also calculate the mean age at childbearing ($\mu$). Remember that
$$ TFR = \sum {_nF_x} \cdot n$$
$$ GRR = TFR \cdot f_{fab}$$
$$NRR = \frac{\sum {_nF_x}\cdot {_nL_x} \cdot f_{fab}}{l_0} $$
$$
\mu = \frac{\sum (x+2.5)\cdot {_nF_x}\cdot {_nL_x}}{\sum \cdot {_nF_x}\cdot {_nL_x}}
$$

```{r}
# calculate TFR, GRR and NRR by country
# set fraction female at birth to be 0.4886
ffab <- 0.4486

d %>% 
  filter(age>=15, age < 50) %>% # restrict our dataset to be fertility rates for women aged 15-49
  group_by(country) %>% 
  summarise(tfr = sum(nFx*5), 
            grr = tfr*ffab,
            nrr = sum(nFx*nLx*ffab/radix),
            mu = sum((age+2.5)*nFx*nLx)/sum(nFx*nLx)) %>%
  arrange(tfr)
```

We can also plot the age-specific fertility rates for each country. 

```{r}
# create a dataset which only has reproductive ages (15-49)
d_rpa <- d %>% filter(age >= 15, age < 50)

# plot age-specific fertility rates
ggplot(data = d_rpa, aes(x = age, y = nFx, color = country)) + 
  geom_line()+
  ylab("Age-specific fertility rate (births/woman)")+
  ggtitle("Age-specific fertility rate by country")+ 
  scale_color_brewer(palette="Set1") + 
  theme_minimal(base_size = 12) ## change ggplot default theme
```

We can also plot the proportion of the population in each age group. This is like half of a population pyramid (the female half). First we need to calculate the proportion in each age group from the raw numbers.

```{r}
# let's calculate the proportion in each age group so we can compare across countries
d <- d %>% 
  group_by(country) %>%
  mutate(nCx = nKx/sum(nKx))
```

```{r}
ggplot(d, aes(x = age, y = nCx)) + 
  facet_wrap(~country, ncol=3)+
  geom_bar(stat="identity", position = "dodge")+
  ggtitle("Proportion of population in each age group")+
  ylab("proportion")+
  coord_flip()
```

## Section II: Population Projections

In this section, we will do a population projection exercise using the fertility, mortality and population data from the dataset from the first section of this lab. We will then compare these results to the growth rate and population structure implied by stable population theory. 

We will use the same data from the earlier exercises. 

### Leslie matrices

We need to create a Leslie matrix which contains the fertility and survivorship information in order to project a population forward. Elements in the top line of the Leslie matrix are equal to 
$$
_nL_0 \cdot \frac{1}{2}(_nF_x + {_nF_{x+n}}\cdot\frac{_nL_{x+n}}{_nL_x}) \cdot F_{fab}
$$
and the subdiagonals are equal to
$$
\frac{_nL_{x+n}}{_nL_x}
$$
The last interval, because it is open-ended, is equal to
$$
\frac{Tx}{Tx_{-n}}
$$

for all groups except the open-ended age group (see lecture slides for how we get the bottom row of the Leslie matrix).

Below is a function that helps to create a Leslie matrix. It takes four inputs: 

- a vector of $_nL_x$ values 
- a vector of $_nF_x$ values
- the number of age groups is 18, because we are considering the five year age groups from ages 0--85+.
- The fraction female at birth, which is set to a default value of 0.4886. 

With these inputs, the function `leslie` creates a square matrix with dimension equal to `n_age_groups` and fills in the top row and subdiagonal according to the equations above. 

```{r}
## function to define Leslie matrix, based on nLx and nFx values
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=18,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  # last open-ended age group
  L[n_age_groups, n_age_groups-1] =   L[n_age_groups, n_age_groups] = nLx[n_age_groups] / (nLx[n_age_groups] + nLx[n_age_groups - 1]) 
  L[1, n_age_groups] = 0

  return(L)
}
```

We can use this function to create a Leslie matrix using the $_nL_x$ and $_nF_x$ values for United States:

```{r}
LAA <- leslie(nLx = d$nLx[d$country=="United States"], 
              nFx = d$nFx[d$country=="United States"])
```

Below are Leslie Matrixes for S Korea (LAS), Kenya (LAK) and Niger (LAN)

```{r}
LAS <- leslie(nLx = d$nLx[d$country=="S Korea"], 
              nFx = d$nFx[d$country=="S Korea"])
```

```{r}
LAK <- leslie(nLx = d$nLx[d$country=="Kenya"], 
              nFx = d$nFx[d$country=="Kenya"])
```

```{r}
LAN <- leslie(nLx = d$nLx[d$country=="Niger"], 
              nFx = d$nFx[d$country=="Niger"])
```


Projections

```{r}
n_age_groups <- 18 # 0-85+ in 5yr age groups
n_projections <- 50 # want to project forward 50*5 = 250 years
initial_pop <- d$nKx[d$country=="United States"] # set initial population to be US's population in 2020
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- initial_pop[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- LAA%*%K[,i-1] 
}

head(K[,1:5])
```

```{r}
#add 100 immigrant to rows 4-6 (300,000 people)
K5 <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K5[,1] <- initial_pop[1:n_age_groups]

#projection with immigration add 100 to rows 4-6 here at each projection step
for(i in 2:(n_projections+1)){
  K5[,i] <- LAA%*%K5[c(4:6) <- c(100,100,100)][,i-1]
}

head(K5[,1:5])
```

South Korea's rates to the United States.

```{r}

K2 <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K2[,1] <- initial_pop[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K2[,i] <- LAS%*%K2[,i-1] 
}

head(K2[,1:5])
```

Kenya

```{r}
K3 <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K3[,1] <- initial_pop[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K3[,i] <- LAK%*%K3[,i-1] 
}

head(K3[,1:5])
```
Niger

```{r}
K4 <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K4[,1] <- initial_pop[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K4[,i] <- LAN%*%K4[,i-1] 
}

head(K4[,1:5])
```

move into dataframes

```{r}
# get K in a form that's easier to ggplot
# make into a dataframe, name columns as years and add an age column
Kdf <- as.data.frame(K)
colnames(Kdf) <- seq(from = 2020, to = (2020+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 85, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -age) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         age = as.factor(age)) %>%
  mutate(totalpop = sum(population[age == 0]))

head(Klong)
```
K2 (SK)
```{r}
Kdf2 <- as.data.frame(K2)
colnames(Kdf2) <- seq(from = 2020, to = (2020+n_projections*5), by = 5)
Kdf2 <- cbind(age = seq(from = 0, to = 85, by = 5), Kdf2)

# get in long format and then add proportion of population in each age group
Klong2 <- Kdf2 %>% 
  gather(year, population, -age) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         age = as.factor(age))
head(Klong2)
```
K3 (Kenya)
```{r}
Kdf3 <- as.data.frame(K3)
colnames(Kdf3) <- seq(from = 2020, to = (2020+n_projections*5), by = 5)
Kdf3 <- cbind(age = seq(from = 0, to = 85, by = 5), Kdf3)

# get in long format and then add proportion of population in each age group
Klong3 <- Kdf3 %>% 
  gather(year, population, -age) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         age = as.factor(age))
head(Klong3)
```
K4(Niger)

```{r}
Kdf4 <- as.data.frame(K4)
colnames(Kdf4) <- seq(from = 2020, to = (2020+n_projections*5), by = 5)
Kdf4 <- cbind(age = seq(from = 0, to = 85, by = 5), Kdf4)

# get in long format and then add proportion of population in each age group
Klong4 <- Kdf4 %>% 
  gather(year, population, -age) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         age = as.factor(age)) 
         
head(Klong4)

Kdf4_wapop <- c(year = 1, colSums(Kdf4[c(4:13),])[-1])
Kdf4_wapop
depratio <- colSums(Kdf4[c(1:3, 14:18),])/Kdf4_wapop
plot(depratio[-1])


```


Create a data frame that has the total population by year, and calculate the annual growth rate based on the equation

$$
K(t+5) = K(t) \cdot e^{5R}
$$
So
$$
R = \frac{1}{5} \cdot \log \frac{K(t+5)}{K(t)}
$$

```{r}
# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5)) 
head(tot_pop)
```

```{r}
Dependecy.US <- Kdf %>% 
  summarise(Kdf_wapop=c(year = 1, colSums(Kdf[c(4:13),])[-1])) %>%
  mutate(depratio = (colSums(Kdf[c(1:3, 14:18),])/Kdf_wapop))

Dependecy.US = Dependecy.US[-1,]
Dependecy.US = Dependecy.US[-1]

Dependecy.US = Dependecy.US[,1]

Dependecy.US

tot_pop = cbind(tot_pop, Dependecy.US)
```

Below is the calculation of dependency ratio for Kenya rates, added into the total pop dataframe
```{r}
Dependecy.SK <- Kdf2 %>% 
  summarise(Kdf2_wapop=c(year = 1, colSums(Kdf2[c(4:13),])[-1])) %>%
  mutate(depratio = (colSums(Kdf2[c(1:3, 14:18),])/Kdf2_wapop))

Dependecy.SK = Dependecy.SK[-1,]
Dependecy.SK = Dependecy.SK[-1]

Dependecy.SK = Dependecy.SK[,1]

Dependecy.SK

tot_pop2 = cbind(tot_pop2, Dependecy.SK)
```


Below is the calculation of dependency ratio for Kenya rates, added into the total pop dataframe
```{r}
Dependecy.Kenya <- Kdf3 %>% 
  summarise(Kdf3_wapop=c(year = 1, colSums(Kdf3[c(4:13),])[-1])) %>%
  mutate(depratio = (colSums(Kdf3[c(1:3, 14:18),])/Kdf3_wapop))

Dependecy.Kenya = Dependecy.Kenya[-1,]
Dependecy.Kenya = Dependecy.Kenya[-1]

Dependecy.Kenya = Dependecy.Kenya[,1]

Dependecy.Kenya

tot_pop3 = cbind(tot_pop3, Dependecy.Kenya)
```

Below is the calculation of dependency ratio for Niger rates, added into the total pop dataframe
```{r}
Dependecy.Niger <- Kdf4 %>% 
  summarise(Kdf4_wapop=c(year = 1, colSums(Kdf4[c(4:13),])[-1])) %>%
  mutate(depratio = (colSums(Kdf4[c(1:3, 14:18),])/Kdf4_wapop))

Dependecy.Niger = Dependecy.Niger[-1,]
Dependecy.Niger = Dependecy.Niger[-1]

Dependecy.Niger = Dependecy.Niger[,1]

Dependecy.Niger

tot_pop4 = cbind(tot_pop4, Dependecy.Niger)
```


For South Korea

```{r}
tot_pop2 <- Klong2 %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))
head(tot_pop2)
```
For Kenya

```{r}
tot_pop3 <- Klong3 %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))
head(tot_pop3)
```
For Niger

```{r}
tot_pop4 <- Klong4 %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))
head(tot_pop4)
```

Now we can plot total population, growth rate, and proportion in each age group over time.  

```{r}

# plot total population over time 
ggplot(data = tot_pop[1,2], aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Total population over time")

# plot growth rate over time
ggplot(data = tot_pop, aes(x = year, y = R, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Growth rate of population over time")

# plot proportion in each age group over time
ggplot(data = Klong, aes(x = year, y = proportion, group = age, color = age)) +
  geom_point() + geom_line()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Proportion of population in each age group over time")

```

